---
title: "Kmer_GWAS_filter_loci"
author: "Cécile Lorrain"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    keep_md: true
---

```{r setup, message = F, warning = F}
library(tximportData)
library(tximport)
library(tidyverse)
library(ape)
library(RColorBrewer)
library(pheatmap)
library(ggpubr)
library(WGCNA)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```


# Kmer fitlering reference-free

### Step 0: Concatentate the significant kmers per phenotype, this will give the total number of significant kmers (repeated, multi-mapped or unique)

```{bash, eval = F}
# Define the output file
output_file="/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/step0_allkmers_pass_threshold_10per_for_fetch.blast.tsv"

# Clear the output file if it already exists
> "$output_file"

# Set a flag for including headers
header_included=false

# Loop over each directory and concatenate files
for dir in /Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/only_signif_phenotypes/Silvias_GWAS_output_dir_*/kmers/; do
    # Get the file name
    file="${dir}/pass_threshold_10per"
    
    # Get the directory name to add as a column (basename will extract just the directory part)
    dir_name=$(basename "$(dirname "$dir")")
    
    # Check if the file exists
    if [[ -f "$file" ]]; then
        # If headers haven't been included, add the header from the first file and add a new "Source" column
        if [[ "$header_included" == false ]]; then
            # Add the header line with "Source" as the new column
            awk 'NR==1 {print $0 "\tSource"}' "$file" >> "$output_file"
            header_included=true
        fi

        # Add data rows, appending the directory name to each line, skipping the header
        awk -v source="$dir_name" 'NR>1 {print $0 "\t" source}' "$file" >> "$output_file"
    fi
done

echo "Concatenation complete. Output saved to $output_file."

```

This gives us the total number of significant kmers throughout all phenotypes that pass the 10 percent threshold. This file contain all kmers so there is some redundancy. We will now filter the unique kmers; meaning the kmers sequences.

### Step 1: List the "unique" kmers
  
  The kmers-gwas analysis attributes a kmer-id as "rs" per GWAS, meaning that the same kmers can be found in different GWAS.
  The first step is list the different unique kmers based on their sequence. For this, I used the following code:
  
```{r}
# Read the concatenated file
all_kmers <- read.delim("/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/step0_allkmers_pass_threshold_10per_for_fetch.blast.tsv")

# Add a new column 'phenotype' by extracting everything after 'Silvias_GWAS_output_dir_' in the 'Source' column
all_kmers$phenotype <- sub(".*Silvias_GWAS_output_dir_", "", all_kmers$Source)

# Split the 'rs' column into two columns, 'rs_sequence' and 'rs_number'
all_kmers <- transform(all_kmers, 
                       rs_sequence = sub("_.*", "", rs), 
                       rs_number = as.numeric(sub(".*_", "", rs)))

# Display the first few rows to check the result
head(all_kmers)

# Filter for unique rows based on 'rs_sequence' while keeping all columns
unique_kmers <- all_kmers %>% 
  distinct(rs_sequence, .keep_all = TRUE)%>%
  arrange(rs_sequence)

# Add a unique ID column with format "kmer0001", "kmer0002", etc.
unique_kmers <- unique_kmers %>%
  dplyr::mutate(kmer_shortID = sprintf("kmer%04d", row_number()))
write.table(unique_kmers, "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/step1_unique_kmers.tsv", sep = "\t", row.names = FALSE)


#let's add the temperature data:
raw_signif_kmers <- left_join(all_kmers, unique_kmers[,c(12,14)]) %>%
  mutate(temp = case_when(
    grepl("18C", phenotype) ~ 18,
    grepl("30C", phenotype) ~ 30,
    grepl("0C", phenotype) ~ 0,
    TRUE ~ NA_real_  # Use NA if no match is found
  ))
```

## Step 2: Filter the kmers that are found in the control conditions 
  
  We can put the "control" aside but it could be interesting to see if some specific kmers are found in the control conditions.

```{r}
kmers_common_control_or_only_in_control <- raw_signif_kmers %>%
  # Group by kmer_shortID to check conditions within each group
  dplyr::group_by(kmer_shortID) %>%
  # Create flags for common control or only in 18
  dplyr::mutate(
    # Check if any temp == 18 and any of temp == 0 or temp == 30
    common_control = any(temp == 18) & any(temp == 0 | temp == 30),
    # Check if all occurrences of kmer_shortID are only in temp == 18
    only_in_18 = all(temp == 18)
  ) %>%
  # Filter for kmer_shortIDs that are either common_control or only_in_18
  dplyr::filter(common_control | only_in_18) %>%
  # Ungroup the data to remove grouping
  ungroup()

kmers_only_in_shocks <- anti_join(raw_signif_kmers, kmers_common_control_or_only_in_control)

#count kmers per phenotype
kmers_per_phenotype <- raw_signif_kmers %>%
  dplyr::group_by(phenotype) %>%
  dplyr::summarise(num_kmers_raw = n_distinct(kmer_shortID))
kmers_per_phenotype1 <- kmers_only_in_shocks %>%
  dplyr::group_by(phenotype) %>%
  dplyr::summarise(num_kmers_only_in_shocks = n_distinct(kmer_shortID))
kmers_per_phenotype <- left_join(kmers_per_phenotype, kmers_per_phenotype1)
```


## Step 3: Extract the genes from gff and concatenate into a pagenome_annotation" gff file

 I will add the pangenome.gff and the orthogroups tables in the drive so you don't have to re-create it.

```{bash, eval = F}
# Extract the genes from GFF
for file in *.gff3; do
  awk '$3 == "mRNA" || $3 == "CDS" {print $0}' "$file" > "${file%.gff3}_mRNA_CDS.gff3"
done

# Concatenate the GFF files
cat *_mRNA_CDS.gff3 > pangenome_annotation.gff3

# Convert GFF to BED format with columns: 1=seqname, 4=start, 5=end, 7=strand, 9=attributes
awk -v OFS="\t" '$3 == "mRNA" || $3 == "CDS" {print $1, $4-1, $5, ".", ".", $7}' pangenome_annotation.gff3 > all_genes_pangenome.bed

# Sort the BED file
sort -k1,1 -k2,2n all_genes_pangenome.bed > all_genes_pangenome_sorted.bed

# Standardize chromosome names if necessary
sed 's/\(chr_\)\([1-9]\)\([[:space:]]\)/\1\0\2\3/' all_genes_pangenome_sorted.bed > all_genes_pangenome_sorted_standardized.bed

```

## Step 5: Concatenate the Kmer GWAS filter loci after mapping onto the 19 pangenome references

```{python}
import os
import pandas as pd
import glob

# Base directory containing all target directories
base_dir = "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/only_signif_phenotypes/"

# Strains to exclude
exclude_strains = ["08STCH015", "08STCZ015", "08STF040", "09STD078"]

# Get list of directories starting with "Silvias_GWAS_output_dir_"
directories = [d for d in os.listdir(base_dir) if os.path.isdir(os.path.join(base_dir, d)) and d.startswith("Silvias_GWAS_output_dir_")]

# Loop over each directory and concatenate files
for directory in directories:
    dir_path = os.path.join(base_dir, directory)
    output_file = os.path.join(base_dir, f"allrefs_significant_kmers10perc_{directory}.tsv")
    
    # Find all significant_kmers_blast_*.tsv files in the directory, excluding specified strains
    files_to_concat = []
    for file in glob.glob(os.path.join(dir_path, "significant_kmers_blast_*.tsv")):
        # Check if file contains any excluded strain names
        if not any(exclude_strain in file for exclude_strain in exclude_strains):
            files_to_concat.append(file)
    
    # Concatenate files if any files remain after filtering
    if files_to_concat:
        # Read all files into a list of dataframes and concatenate
        dfs = [pd.read_csv(file, sep='\t') for file in files_to_concat]
        combined_df = pd.concat(dfs, ignore_index=True)
        
        # Save to output file
        combined_df.to_csv(output_file, sep='\t', index=False)
        print(f"Concatenated {len(files_to_concat)} files into {output_file}")
    else:
        print(f"No eligible files found in {directory} after filtering out excluded strains.")
```

## Step 6: Extract the significant, unique kmers FOR THE SHOCKS (delete the empty files before running)

```{r}
# Define the input directory
input_dir <- "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/only_signif_phenotype/"

# List files that match the pattern with 0C, 18C, or 30C in their names
files <- list.files(input_dir, pattern = "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_*\\.tsv$", full.names = TRUE)
```

```{r}
library(dplyr)
library(stringr)
### Test with one file
# Path to input directory and output directory
input_dir <- "~/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/only_signif_phenotypes/"
output_dir <- "~/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/"

# Specific file for testing
file <- "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_GrowthRate0C.tsv"

# Generate dynamic variable name based on the file name
file_shortname <- str_extract(basename(file), "(?<=allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_).*")

# Read data
gwas_data <- read.table(file.path(input_dir, file), header = TRUE, sep = "\t")

# Process data
gwas_data <- gwas_data %>%
  dplyr::mutate(genome = str_extract(chromosome, "^[^.]+")) %>%
  dplyr::left_join(kmers_only_in_shocks[,c(2,10,11,12,14,15)]) %>%  # Adjust join column if needed
  na.omit() %>%
  dplyr::mutate(position_end = position + 31)

# Uniquely mapped kmers per reference
counts <- gwas_data %>%
  dplyr::group_by(kmer_shortID, genome) %>%
  dplyr::tally(name = "n_kmer") %>%
  dplyr::filter(n_kmer < 2)

gwas_data_unique_map <- gwas_data %>%
  dplyr::left_join(counts, by = c("kmer_shortID", "genome")) %>%
  na.omit()

# Count kmers per chromosome
counts_data <- gwas_data_unique_map %>%
  dplyr::group_by(chromosome) %>%
  dplyr::tally(name = "n_kmer_total") %>%
  dplyr::filter(n_kmer_total > 1)

gwas_data_unique_map <- gwas_data_unique_map %>%
  dplyr::left_join(counts_data) %>%
  na.omit()

# Ensure the data is sorted by chromosome and position
gwas_data_unique_map <- gwas_data_unique_map %>%
  arrange(chromosome, position)

# Identify gaps between consecutive positions
counts_window <- gwas_data_unique_map %>%
  dplyr::group_by(chromosome) %>%
  # Create columns to check gaps between consecutive rows (forward and backward)
  dplyr::mutate(
    next_position = lead(position),  # Get the position of the next row
    gap = next_position - position_end,  # Calculate the gap between the current position_end and the next position
    prev_position_end = lag(position_end),  # Get the position_end of the previous row
    gap2 = position - prev_position_end,  # Calculate the gap between the current position and the previous position_end
    new_window = ifelse(gap > 10000 | gap2 > 10000 | is.na(gap) | is.na(gap2), 1, 0)  # Flag new window if either gap is greater than 10000 or if it's the last row
  ) %>%
  # Create a cumulative window ID based on the new_window flag
  dplyr::mutate(
    window_id = cumsum(new_window) + 1  # Assign a unique ID to each window, starting at 1
  ) %>%
  # Recalculate position_min and position_max within each window
  dplyr::group_by(chromosome, window_id) %>%
  dplyr::mutate(
    position_min = min(position),
    position_max = max(position_end),
    window_size = position_max - position_min
  ) %>%
  # Ungroup after processing
  dplyr::ungroup()

# Write output
output_file <- file.path(output_dir, paste0(file_shortname, "_gwas_kmers.txt"))
write.table(counts_window[c("rs","sseqid","position","chr","ps","n_miss","allele1","allele0","af","l_mle","p_lrt","log10_pval","chromosome","genome","Source","phenotype","rs_sequence","kmer_shortID","temp","position_end","n_kmer_total","position_min","position_max","window_size")], output_file, sep = "\t", row.names = FALSE)

# Print output message
print(paste("Processed and saved:", output_file))

```

```{r, message=F, warning=F}
# Path to input directory and output directory
input_dir <- "~/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/only_signif_phenotypes/"
output_dir <- "~/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/"

# List of files
files <- c(
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_GrowthRate0C.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Interceptx0C.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Interceptx30C.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Mean_grey0C_08dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Mean_grey0C_19dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Mean_grey30C_08dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius0C_08dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius0C_15dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius30C_08dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius30C_12dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Ratio0C.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_SporespermL30C.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Std_grey0C_08dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Std_grey0C_15dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Std_grey0C_19dpi.tsv",
  "allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Std_grey30C_08dpi.tsv"
)

# Function to process each file
process_file <- function(file) {
  # Generate dynamic variable name based on the file name
  file_shortname <- str_extract(basename(file), "(?<=allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_).*")

  # Read data
  gwas_data <- read.table(file.path(input_dir, file), header = TRUE, sep = "\t")

  # Process data
  gwas_data <- gwas_data %>%
    dplyr::mutate(genome = str_extract(chromosome, "^[^.]+")) %>%
    dplyr::left_join(kmers_only_in_shocks[,c(2,10,11,12,14,15)]) %>%  # Adjust join column if needed
    na.omit() %>%
    dplyr::mutate(position_end = position + 31)

  # Uniquely mapped kmers per reference
  counts <- gwas_data %>%
    dplyr::group_by(kmer_shortID, genome) %>%
    dplyr::tally(name = "n_kmer") %>%
    dplyr::filter(n_kmer < 2)

  gwas_data_unique_map <- gwas_data %>%
    dplyr::left_join(counts, by = c("kmer_shortID", "genome")) %>%
    na.omit()

  # Count kmers per chromosome
  counts_data <- gwas_data_unique_map %>%
    dplyr::group_by(chromosome) %>%
    dplyr::tally(name = "n_kmer_total") %>%
    dplyr::filter(n_kmer_total > 1)

  gwas_data_unique_map <- gwas_data_unique_map %>%
    dplyr::left_join(counts_data) %>%
    na.omit()

  # Ensure the data is sorted by chromosome and position
  gwas_data_unique_map <- gwas_data_unique_map %>%
    arrange(chromosome, position)

  # Identify gaps between consecutive positions
  counts_window <- gwas_data_unique_map %>%
    dplyr::group_by(chromosome) %>%
    # Create columns to check gaps between consecutive rows (forward and backward)
    dplyr::mutate(
      next_position = lead(position),  # Get the position of the next row
      gap = next_position - position_end,  # Calculate the gap between the current position_end and the next position
      prev_position_end = lag(position_end),  # Get the position_end of the previous row
      gap2 = position - prev_position_end,  # Calculate the gap between the current position and the previous position_end
      new_window = ifelse(gap > 10000 | gap2 > 10000 | is.na(gap) | is.na(gap2), 1, 0)  # Flag new window if either gap is greater than 10000 or if it's the last row
    ) %>%
    # Create a cumulative window ID based on the new_window flag
    dplyr::mutate(
      window_id = cumsum(new_window) + 1  # Assign a unique ID to each window, starting at 1
    ) %>%
    # Recalculate position_min and position_max within each window
    dplyr::group_by(chromosome, window_id) %>%
    dplyr::mutate(
      position_min = min(position),
      position_max = max(position_end),
      window_size = position_max - position_min
    ) %>%
    # Ungroup after processing
    dplyr::ungroup()

  # Write output
  output_file <- file.path(output_dir, paste0(file_shortname, "_gwas_kmers.txt"))
  write.table(counts_window[c("rs", "sseqid", "position", "chr", "ps", "n_miss", "allele1", "allele0", "af", 
                             "l_mle", "p_lrt", "log10_pval", "chromosome", "genome", "Source", "phenotype", 
                             "rs_sequence", "kmer_shortID", "temp", "position_end", "n_kmer_total", "position_min", 
                             "position_max", "window_size")], output_file, sep = "\t", row.names = FALSE)

  # Print output message
  print(paste("Processed and saved:", output_file))
}

# Apply the function to each file in the list
purrr::walk(files, process_file)

```

## Step 7: Extract the significant, unique kmers FOR THE CONTROL

```{r, message=F, warning=F}
# Path to input directory and output directory
input_dir <- "~/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/only_signif_phenotypes/"
output_dir <- "~/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/"

# List of files
files <- c("allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_GrowthRate18C.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Std_grey18C_08dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Std_grey18C_12dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Std_grey18C_19dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_SporespermL18C.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius18C_08dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius18C_12dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius18C_15dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Radius18C_19dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Mean_grey18C_19dpi.tsv",
"allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_Mean_grey18C_08dpi.tsv")

# Function to process each file
process_file <- function(file) {
  # Generate dynamic variable name based on the file name
  file_shortname <- str_extract(basename(file), "(?<=allrefs_significant_kmers10perc_Silvias_GWAS_output_dir_).*")

  # Read data
  gwas_data <- read.table(file.path(input_dir, file), header = TRUE, sep = "\t")

  # Process data
  gwas_data <- gwas_data %>%
    dplyr::mutate(genome = str_extract(chromosome, "^[^.]+")) %>%
    dplyr::left_join(kmers_common_control_or_only_in_control[,c(2,10,11,12,14,15)]) %>%  # Adjust join column if needed
    na.omit() %>%
    dplyr::mutate(position_end = position + 31)

  # Uniquely mapped kmers per reference
  counts <- gwas_data %>%
    dplyr::group_by(kmer_shortID, genome) %>%
    dplyr::tally(name = "n_kmer") %>%
    dplyr::filter(n_kmer < 2)

  gwas_data_unique_map <- gwas_data %>%
    dplyr::left_join(counts, by = c("kmer_shortID", "genome")) %>%
    na.omit()

  # Count kmers per chromosome
  counts_data <- gwas_data_unique_map %>%
    dplyr::group_by(chromosome) %>%
    dplyr::tally(name = "n_kmer_total") %>%
    dplyr::filter(n_kmer_total > 1)

  gwas_data_unique_map <- gwas_data_unique_map %>%
    dplyr::left_join(counts_data) %>%
    na.omit()

  # Ensure the data is sorted by chromosome and position
  gwas_data_unique_map <- gwas_data_unique_map %>%
    arrange(chromosome, position)

  # Identify gaps between consecutive positions
  counts_window <- gwas_data_unique_map %>%
    dplyr::group_by(chromosome) %>%
    # Create columns to check gaps between consecutive rows (forward and backward)
    dplyr::mutate(
      next_position = lead(position),  # Get the position of the next row
      gap = next_position - position_end,  # Calculate the gap between the current position_end and the next position
      prev_position_end = lag(position_end),  # Get the position_end of the previous row
      gap2 = position - prev_position_end,  # Calculate the gap between the current position and the previous position_end
      new_window = ifelse(gap > 10000 | gap2 > 10000 | is.na(gap) | is.na(gap2), 1, 0)  # Flag new window if either gap is greater than 10000 or if it's the last row
    ) %>%
    # Create a cumulative window ID based on the new_window flag
    dplyr::mutate(
      window_id = cumsum(new_window) + 1  # Assign a unique ID to each window, starting at 1
    ) %>%
    # Recalculate position_min and position_max within each window
    dplyr::group_by(chromosome, window_id) %>%
    dplyr::mutate(
      position_min = min(position),
      position_max = max(position_end),
      window_size = position_max - position_min
    ) %>%
    # Ungroup after processing
    dplyr::ungroup()

  # Write output
  output_file <- file.path(output_dir, paste0(file_shortname, "_gwas_kmers.txt"))
  write.table(counts_window[c("rs", "sseqid", "position", "chr", "ps", "n_miss", "allele1", "allele0", "af", 
                             "l_mle", "p_lrt", "log10_pval", "chromosome", "genome", "Source", "phenotype", 
                             "rs_sequence", "kmer_shortID", "temp", "position_end", "n_kmer_total", "position_min", 
                             "position_max", "window_size")], output_file, sep = "\t", row.names = FALSE)

  # Print output message
  print(paste("Processed and saved:", output_file))
}

# Apply the function to each file in the list
purrr::walk(files, process_file)
```

  From this part, we have the significant, unique kmers for the control and the shocks. We can now merge the kmers with the orthogroups and identify the cross-references loci. 


## Step 8: Convert to bed and use bedtools closest to find the closest gene and TE

```{bash, eval = F}
#!/bin/bash

# List of input files
files=(
"Std_grey18C_19dpi.tsv_gwas_kmers.txt"
"Interceptx0C.tsv_gwas_kmers.txt"
"Interceptx30C.tsv_gwas_kmers.txt"
"Mean_grey0C_19dpi.tsv_gwas_kmers.txt"
"Mean_grey18C_08dpi.tsv_gwas_kmers.txt"
"Mean_grey18C_19dpi.tsv_gwas_kmers.txt"
"Mean_grey30C_08dpi.tsv_gwas_kmers.txt"
"Radius0C_08dpi.tsv_gwas_kmers.txt"
"Radius0C_15dpi.tsv_gwas_kmers.txt"
"Radius18C_08dpi.tsv_gwas_kmers.txt"
"Radius18C_15dpi.tsv_gwas_kmers.txt"
"Ratio0C.tsv_gwas_kmers.txt"
"SporespermL18C.tsv_gwas_kmers.txt"
"SporespermL30C.tsv_gwas_kmers.txt"
"Std_grey0C_08dpi.tsv_gwas_kmers.txt"
"Std_grey18C_08dpi.tsv_gwas_kmers.txt"
"Std_grey18C_12dpi.tsv_gwas_kmers.txt"
)

# Loop through each file
for file in "${files[@]}"; do
  # Define output filename
  output_file="${file%.txt}.bed"

  # Use awk to extract columns, remove quotes, and omit the header
  awk 'NR > 1 {
    gsub(/"/, "", $13);
    gsub(/"/, "", $22);
    gsub(/"/, "", $23);
    gsub(/"/, "", $18);
    gsub(/"/, "", $24);
    print $13, $22, $23, $18, $24
  }' OFS="\t" "$file" > "$output_file"

  echo "Processed and created: $output_file"
done
```

```{bash, eval = F}
#!/bin/bash

# Directory paths
#sed 's/\(chr_\)\([1-9]\)\([[:space:]]\)/\1\0\2\3/' all_genes_pangenome_sorted_corrected.bed > all_genes_pangenome_sorted_standardized.bed
input_dir="/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/"     # Update this with your .bed files' directory path
output_dir="/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/"            # Update with your desired output directory path
orthologs_file="/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/ProtOthro/final_long_table_proteinOrtho.tsv"

# Loop over each .bed file
for bedfile in *.bed 
do
  # Define filenames
  base_name=$(basename "$bedfile" .bed)
  sorted_bed="$output_dir/${base_name}_sorted.bed"
  closest_bed="$output_dir/${base_name}_closest_genes.bed"
  merged_output="$output_dir/${base_name}_merged_with_orthologs.bed"

  # Sort bed file
  sort -k1,1 -k2,2n "$input_dir/$bedfile" > "$sorted_bed"

  # Run bedtools closest with filtering for distance within 1kb
  bedtools closest -a "$sorted_bed" -b /Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/bed_files/all_genes_pangenome_standardized_sorted.bed -d > "$closest_bed"

  echo "Processed and merged: $bedfile"
done
```

```{bash, eval = F}
#!/bin/bash

# Directory containing your *.tsv_gwas_kmers_sorted.bed files
input_dir="/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/" 

# Reference BED file for bedtools closest
reference_bed="/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/ref_annotations/all_pangenome_TE_2023_standardized_sorted.bed"

# Output directory (optional: set to the same as input_dir if no separate output directory is needed)
output_dir="$input_dir"  # Adjust this if you want a different output directory

# Loop through each .tsv_gwas_kmers_sorted.bed file in the directory
for bed_file in "$input_dir"/*.tsv_gwas_kmers_sorted.bed; do
    # Extract the filename without extension
    filename=$(basename "$bed_file" .tsv_gwas_kmers_sorted.bed)
    
    # Define the output file path
    output_file="$output_dir/${filename}_TE_closest.bed"
    
    # Run bedtools closest and filter for distance == 0
    bedtools closest -a "$bed_file" -b "$reference_bed" -d | awk '$NF == 0' > "$output_file"
    
    echo "Processed $bed_file and saved overlapping TEs to $output_file"
done

echo "All files processed!"

```

  Now wa have the closest genes and TEs for each kmer. We can now merge with orthogroups and identify cross-references loci.

## Step 9: Merge with orthogrroups and identify cross-refs loci

"Common Region": If any overlapping kmer windows (even with different kmer_shortIDs) share the same orthogroup(s), it will be classified as "common." This requires checking for overlapping regions and determining if they share the same orthogroups.

"Specific Region": If there is no overlap or the orthogroups are unique to each kmer_shortID, it will be classified as "specific."

  After that I still re-examined the files manually to curate the loci. 


```{r}
library(dplyr)
library(tidyr)

# Define input and output directories, column headers, and path to the ortho data file
input_dir <- "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/gene_closest_bed/"
output_dir <- "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/combined_output/"
ortho_file <- "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/ProtOthro/final_long_table_proteinOrtho.tsv"

# Define column headers for the bed data
headers <- c("chromosome", "position_min", "position_max", "kmer_shortID", "window_size",
             "chromosome_gff", "start", "end", "strand", "geneID", "distance_sig_kmers")

# Load Ortho data file and select relevant columns
ortho_data <- read.table(ortho_file, header = TRUE, sep = "\t")
ortho_data <- ortho_data[, c("geneID", "OrthoID")]

# List of files to process
files <- list(
"Std_grey18C_19dpi.tsv_gwas_kmers_closest_genes.bed",
"Std_grey18C_12dpi.tsv_gwas_kmers_closest_genes.bed",
"Std_grey18C_08dpi.tsv_gwas_kmers_closest_genes.bed",
"Std_grey0C_08dpi.tsv_gwas_kmers_closest_genes.bed",
"SporespermL30C.tsv_gwas_kmers_closest_genes.bed",
"SporespermL18C.tsv_gwas_kmers_closest_genes.bed",
"Radius18C_15dpi.tsv_gwas_kmers_closest_genes.bed",
"Ratio0C.tsv_gwas_kmers_closest_genes.bed",
"Radius18C_08dpi.tsv_gwas_kmers_closest_genes.bed",
"Radius0C_15dpi.tsv_gwas_kmers_closest_genes.bed",
"Radius0C_08dpi.tsv_gwas_kmers_closest_genes.bed",
"Mean_grey30C_08dpi.tsv_gwas_kmers_closest_genes.bed",
"Mean_grey18C_19dpi.tsv_gwas_kmers_closest_genes.bed",
"Mean_grey18C_08dpi.tsv_gwas_kmers_closest_genes.bed",
"Mean_grey0C_19dpi.tsv_gwas_kmers_closest_genes.bed",
"Interceptx30C.tsv_gwas_kmers_closest_genes.bed",
"Interceptx0C.tsv_gwas_kmers_closest_genes.bed"
)

# Process each file in the list
for (file in files) {
  file_path <- file.path(input_dir, file)
  if (!file.exists(file_path)) {
    message(paste("File does not exist:", file_path))
    next  # Skip this file if it doesn't exist
  }
  # Continue with processing as before...
}

for (file in files) {
  # Construct the file path and read the .bed file with headers
  file_path <- file.path(input_dir, file)
  bed_data <- read.table(file_path, header = FALSE, sep = "\t", col.names = headers)
  
  # Extract phenotype from file name and add as a new column
  phenotype <- sub("(.*)\\.tsv_gwas_kmers_closest_genes\\.bed", "\\1", file)
  bed_data$phenotype <- phenotype
  
  # Separate the `chromosome_gff` column into `reference` and `CHR`
  bed_data <- bed_data %>%
    separate(chromosome_gff, into = c("reference", "CHR"), sep = "\\.chr_", remove = FALSE)
  
  # Merge bed data with ortho data by geneID
  merged_data <- merge(bed_data, ortho_data, by = "geneID", all.x = TRUE)
  
  # Summarize data by grouping and counting `kmer_shortID`
  merged_data_summarised <- merged_data %>%
    dplyr::group_by(reference, CHR, chromosome, position_min, position_max, kmer_shortID, window_size, geneID, start, end, strand, distance_sig_kmers, phenotype, OrthoID) %>%
    dplyr::summarise(kmer_count = n(), .groups = 'drop')
  
  # Define the output file path and write the summarized data to CSV
  output_file <- file.path(output_dir, paste0(phenotype, "_summarized.csv"))
  write.csv(merged_data_summarised, output_file, row.names = FALSE)
}
```

 Merge with TEs; this is only applied if TEs are found overlapping with the kmers. 


```{r}
# Load necessary libraries
library(dplyr)
library(readr)

# Directories
combined_dir <- "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/combined_output/"
te_dir <- "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/TE_onverelap_bed/"
output_dir <- "/Users/lorrain/Desktop/PROJECTS/5_Silvia_thesis/Chapter2/Kmers/Results_from_Alice/CL_final_tables_shockgwas/output_gwas_loci/new/final_merged_output_if_TEs_found/"

te_headers <- c("chromosome", "position_min", "position_max", "kmer_shortID", "window_size", 
                "chromosome_gff_TE", "TE_start", "TE_end", "TE_ID", "distance_sig_kmers")

# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) dir.create(output_dir)

# List all summarized files with full paths
summarized_files <- file.path(combined_dir, c(
  "Std_grey18C_19dpi_summarized.csv",
  "Std_grey18C_08dpi_summarized.csv",
  "Std_grey0C_08dpi_summarized.csv",
  "Radius18C_15dpi_summarized.csv",
  "Radius18C_08dpi_summarized.csv",
  "Radius0C_08dpi_summarized.csv",
  "Mean_grey30C_08dpi_summarized.csv",
  "Mean_grey18C_19dpi_summarized.csv",
  "Mean_grey18C_08dpi_summarized.csv",
  "Mean_grey0C_19dpi_summarized.csv",
  "Interceptx30C_summarized.csv"
))

# List all TE files with full paths
te_files <- file.path(te_dir, c(
  "Std_grey18C_19dpi_TE_closest.bed",
  "Std_grey18C_08dpi_TE_closest.bed",
  "Std_grey0C_08dpi_TE_closest.bed",
  "Radius18C_15dpi_TE_closest.bed",
  "Radius18C_08dpi_TE_closest.bed",
  "Radius0C_08dpi_TE_closest.bed",
  "Mean_grey30C_08dpi_TE_closest.bed",
  "Mean_grey18C_19dpi_TE_closest.bed",
  "Mean_grey18C_08dpi_TE_closest.bed",
  "Mean_grey0C_19dpi_TE_closest.bed",
  "Interceptx30C_TE_closest.bed"
))

# Iterate over each summarized file and its corresponding TE file
for (summarized_file in summarized_files) {
  
  # Load summarized CSV file
  summarized_df <- read_csv(summarized_file)
  
  # Find corresponding TE file
  te_file <- gsub("_summarized\\.csv", "_TE_closest.bed", basename(summarized_file))
  te_path <- file.path(te_dir, te_file)
  
  if (file.exists(te_path)) {
    # Load TE file with specified headers
    te_df <- read.table(te_path, header = FALSE, sep = "\t", col.names = te_headers, stringsAsFactors = FALSE)
  
    
    # Perform the join based on shared columns
    merged_df <- dplyr::left_join(summarized_df, te_df, by = c("chromosome", "position_min", "position_max", "kmer_shortID", "window_size"))
    
    # Output file path
    output_file <- file.path(output_dir, paste0("final_", basename(summarized_file)))
    
    # Save the merged dataframe
    write_csv(merged_df, output_file)
    
    cat("Merged file saved:", output_file, "\n")
    
  } else {
    cat("TE file not found for:", summarized_file, "\n")
  }
}

cat("All files processed.\n")


```




